---
title: "wild_yeast_growth_curve_example"
output: html_document
date: "2025-07-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("gcplyr", quietly = TRUE)) {
    install.packages("gcplyr")
}
if (!requireNamespace("reshape2", quietly = TRUE)) {
    install.packages("reshape2")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
    install.packages("ggplot2")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
    install.packages("dplyr")
}
library(gcplyr)
library(reshape2)
library(ggplot2)
library(dplyr)
```

## Growth curve analysis

We will be using gcplyr to analyze our growth curves
You can read more about this package here: https://github.com/mikeblazanin/gcplyr?tab=readme-ov-file


## Importing the data

In this section we will import the data. The growth curve data file must be in the same folder/directory on your computer as this R-file. 

```{r data import}
#read in wide-formatted data
all_data<-read_wides(files="combined_curve.tsv")

#transform the wide-data to tidy formatted ata
all_data.tidy<-melt(data = all_data, id.vars = c("file","Time"), value.name = "Measurements", variable.name = "Well")

#read in the file with the metadata
metadata<-read.table(file="strain_sample.txt", header=T)

#convert the data from wide to long format
all_data.tidy<-full_join(all_data.tidy, metadata, relationship = "many-to-many")

#clean up the data and remove the "mistake" well
all_data.tidy$Time<-as.numeric(all_data.tidy$Time)
all_data.tidy$Measurements<-as.numeric(all_data.tidy$Measurements)
all_data.tidy<-subset(all_data.tidy, Well_Type != "mistake")


```

## Pre-processing: subtracting blanks

"Many growth curves are collected by measuring the absorbance or optical density of a culture. However,
with such data an absorbance value of 0 is not equal to a cell density of 0, since components of the media
often absorb some light. It’s best practice to have at least one ‘blank’ well in your plate containing only
media and no cells, so that you can subtract out this difference from your data so that the values you are
working with are scaled correctly.

Here we have some data including a blank well. The first thing you should always do is plot your blank wells
data to ensure they look correct:"


```{r blanks}

#plot the raw data so we can see if the blanks are what we think
ggplot(data = all_data.tidy, aes(x = Time, y = Measurements, color = Well_Type)) + geom_point()

#subtract the "Blank" wells
mean_blank <- mean(dplyr::filter(all_data.tidy, Well_Type == "BLANK")$Measurements)
mean_blank
all_data.tidy$Measurements <- all_data.tidy$Measurements - mean_blank

```


## Pre-processing additional steps

```{r pre process 2}


#Convert time to hours
all_data.tidy$Time <- all_data.tidy$Time/3600 

#Add in some useful metrics/calculations that we will need for the metrics below
all_data.tidy <-
mutate(group_by(all_data.tidy, Well, Sample),
deriv = calc_deriv(x = Time, y = Measurements),
deriv_percap5 = calc_deriv(x = Time, y = Measurements,
percapita = TRUE, blank = 0,
window_width_n = 5, trans_y = "log"),
doub_time = doubling_time(y = deriv_percap5))


```

## Plot all the data for all wells

```{r plot}

#Plot all the data!
ggplot(data = all_data.tidy, aes(x = Time, y = Measurements)) +
geom_line() +
facet_wrap(~Well, nrow = 8, ncol = 12)

```


# Analyze your samples! 

We are going to calculate the following metrics for your samples

Most common metrics
• the lag time
• the maximum cellular growth rate (i.e. minimum doubling time)
• the maximum density (e.g. carrying capacity)
• the area under the curve

## Step 1 

Subset the data to include only your two samples 
I will conduct the tutorial using samples for PAG

```{r subsetting the data}

#Subset the data so it only contains YOUR samples
samples<-subset(all_data.tidy, Sample == "PAG-052825-01-A" | Sample == "PAG-052825-02-A" )

#Look at your data
ggplot(data = samples, aes(x= Time, y=Measurements, color=Sample))+facet_wrap(~Sample)+geom_point()

```
"Bacteria often have a period of time before they reach their maximum growth rate. If you would like to
quantify this lag time, you can use the lag_time function. lag_time needs the x and y values, as well
as the (per-capita) derivative and the blank value (the value of your Measurements that corresponds to a
population density of 0; if your data have already been normalized, simply add blank = 0). It will find the
maximum derivative, then project the tangent line with that slope back until it crosses the starting density."

```{r Lag Time}

#Calculate the lag time
samples.lagtime <-
summarize(group_by(samples, Well, Sample),
lag_time = lag_time(y = Measurements, x = Time,
deriv = deriv_percap5, blank = 0))

#This is the lag time for your samples
samples.lagtime

#Plot your growth curves with the lag time
ggplot(data = samples,
aes(x = Time, y = log(Measurements))) +
geom_point() + facet_wrap(~Well) + geom_vline(data = samples.lagtime, aes(xintercept = lag_time), lty = 2) 

```

# Maximum growth rate and minimum doubling time

"If you want to calculate the bacterial maximum growth rate (i.e. the minimum doubling time), it will often be
sufficient to use max_gc on the per-capita derivatives we calculated in vignette("gc05_process"). (max_gc
works just like R’s built-in max, but with better default settings for growth curve analyses with summarize)."


```{r max growth rate }

#Calculate the maximum growth rate and doubling time
samples.maxgrowth <-
summarize(group_by(samples, Well, Sample),
max_percap = max_gc(deriv_percap5, na.rm = TRUE),
max_percap_time = extr_val(Time, which_max_gc(deriv_percap5)),
doub_time = doubling_time(y = max_percap),max_percap_dens = Measurements[which_max_gc(deriv_percap5)])

#These are the results
samples.maxgrowth

#Plot your data with the growth rates
ggplot(data = samples,
aes(x = Time, y = log(Measurements))) +
geom_point() + facet_wrap(~Well) + geom_vline(data = samples.lagtime, aes(xintercept = lag_time), lty = 2)+
geom_abline(data = samples.maxgrowth, color = "red", aes(slope = max_percap, intercept = log(max_percap_dens) - max_percap*max_percap_time)) 

```



# Maximum density


"The maximum bacterial density can be a measure of bacterial growth yield/efficiency. If your bacteria
plateau in density, the maximum density can also be a measure of bacterial carrying capacity. If you want
to quantify the maximum bacterial density, we can use max_gc to get the global maxima of Measurements
(max_gc, which_max_gc, and extr_val work just like R’s built-in max, which.max, and [, but with better
default settings for growth curve analyses with summarize).
See Peak bacterial density for identifying local maxima of Measurements (e.g. if you wanted the first peak
in Well E11 shown below)."

```{r max density}

#Calculate the maximum density of your sample
samples.maxdens <-
summarize(group_by(samples, Well, Sample),
max_dens = max_gc(Measurements, na.rm = TRUE),
max_time = extr_val(Time, which_max_gc(Measurements)))

#Maximum density
samples.maxdens

#Plot the growth data with the maximum density
ggplot(data = samples, aes(x = Time, y = Measurements)) +
geom_point() + facet_wrap(~Well) + 
  geom_point(data = samples.maxdens, aes(x = max_time, y = max_dens), size = 2, color = "red")

```


# Area under the curve

"The area under the curve is a common metric of total bacterial growth, for instance in the presence of
antagonists like antibiotics or phages. If you want to calculate the area under the curve, you can use the
gcplyr function auc. Simply specify Time as the x and Measurements as the y data whose area-under-thecurve you want to calculate."


```{r auc}

#Calculate the area under the curve
samples.auc <-
summarize(group_by(samples, Sample, Well),
auc = auc(x = Time, y = Measurements))

#Report area under the curve
samples.auc

```

BONUS - make a pretty plot

Take this plot and make it look pretty

```{r pretty plot}

ggplot(data = samples,
aes(x = Time, y = log(Measurements))) +
geom_point() + facet_wrap(~Well) + geom_vline(data = samples.lagtime, aes(xintercept = lag_time), lty = 2)+
geom_abline(data = samples.maxgrowth, color = "red", aes(slope = max_percap, intercept = log(max_percap_dens) - max_percap*max_percap_time)) 
```
